---
title: "Data Mining Final Project"
author: "Arpan Chatterji, Jacob Bulzak, Rajsitee Dhavale"
date: "5/6/2022"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(rpart)
library(rpart.plot)
library(rsample) 
library(randomForest)
library(lubridate)
library(modelr)
library(caret)
library(readr)
library(lattice)
library(caTools)
library(ggplot2)
library(dplyr)
library(data.table)
library(onehot)



```

## Random Forest

We arrived at the conclusion that a random forest model was a consistent winner when it came to performance relative to the other models tested.  After all, random forests are the benchmark standard for supervised learning techniques and have the added benefit of being remarkably easy to implement.

```{r confusion matrix, echo=FALSE}
conf <- startup.forest$confusion
conf
```

The confusion matrix presented above corresponds to 

```{r , echo=FALSE}
plot(startup.forest)

```
The plot above shows out-of-bag MSE as a function of the number of trees used. In each case we see that the number the error initially decreases rapidly at low numbers of trees used but then rapidly decreases and plateaus. Overall, at more than 100 trees there is very little appreciable change in the error.

green: out of bag error

black: RMSE

red:


```{r , echo=FALSE}
modelr::rmse(startup.tree, startup_test)

modelr::rmse(startup.forest, startup_test)

```

In order to ascertain which variables had the strongest effects on whether a startup would be acquired, we introduced the variable importance plot made from our forest model.

```{r , echo=FALSE}
viforest<-varImpPlot(startup.forest, type=1)
```
Discuss variable importance plot!!!


```{r , echo=FALSE}
pd_funding<-partialPlot(startup.forest, startup_test, 'funding_total_usd', las=1)

pd_milestones<-partialPlot(startup.forest, startup_test, 'milestones', las=1)

pd_top500<-partialPlot(startup.forest, startup_test, 'is_top500', las=1)

pd_CA<-partialPlot(startup.forest, startup_test, 'is_CA', las=1)

```

Discuss partial dependence




